Equalizer Programming Guide

### 2009-01-13 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add section describing network layer
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.2 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{Distributed Objects}

@@

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.3 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\subsection{\label{sNetObject}Distributed Objects}

@@

\subsection{\label{sNetwork}The Network Layer}

The Equalizer networking layer provides a peer-to-peer communication
infrastructure. It is used by Equalizer to communicate between the
application node, the server and the render clients. It can also be
used by applications to implement distributed processing independently
or complementary to the core Equalizer functionality.

The network layer is implemented in the \textsf{eq::net} namespace. It
provides layered functionality, which provide higher level functionality
for the programmer. The main primitives in the network layer are:

\begin{description}
\item[Connection] A stream-oriented point-to-point communication
  line. Different implementations of a connection exist.
\item[Node] The abstraction of a process in the cluster. Nodes
  communicate with each other using connections.
\item[Session] Provides unique identifiers and object mapping to a set
  of nodes.
\item[Object] Provides object-oriented data distribution of C++ objects
  between nodes within a session.
\end{description}

\subsubsection{Connections}

The \textsf{eq::net::Connection} is the basic primitive used for
communication between processes in Equalizer. It provides a
stream-oriented communication between two endpoints.

A connection is either closed, connected or listening. A closed
connection can not be used for communications. A connected connection
can be used to read or write data to the communication peer. A listening
connection can accept connection requests.

An \textsf{eq::net::ConnectionSet} is used to manage multiple
connections. The typical use case is to have one or more listening
connections for the local process, and a number of connected connections
for communicating with other processes.

The connection set is used to select one connection which requires some
action. This can be a connection request on a listening connection,
pending data on a connected connection or the notification of a
disconnect.

The connection and connection set can be used by applications to
implement other networking-related functionality, e.g., to communicate
with a sound server on a different machine.

\subsubsection{Packet Handling}

Nodes, sessions and objects communicate using packets derived from
\textsf{eq::net::Packet}. The basic packet dispatch and handling is
implemented in the \textsf{eq::net::Base} class, from which
\textsf{eq::net::Node}, \textsf{eq::net::Session} and
\textsf{eq::net::Object} is subclassed.

The network base class allows the registration of a packets with a
dispatch queue and an invocation method. Each packet has a command
identifier, which is used to identify the registered queue and
method. The method \textsf{dispatchCommand} pushes the packet to the
registered queue, whereas the method \textsf{invokeCommand} calls the
registered command function.

A command function groups the method and \textsf{this} pointer, allowing
to call a C++ method on a concrete instance. If no queue is registered
for a certain command, \textsf{dispatchCommand} directly calls the
registered command function.

This dispatch and invocation functionality is used within Equalizer to
dispatch commands from the receiver thread to the appropriate node or
pipe thread, and then to invoke the command when it is processed by
these threads.

\subsubsection{Nodes}

The \textsf{eq::net::Node} is the abstraction of one process in the
cluster. Each node has a unique identifier, based on the UUID
standard\footnote{\link{http://en.wikipedia.org/wiki/Universally\_Unique\_Identifier}}. This
identifier is used to address nodes, e.g., in order to query connection
information to connect to the node. Nodes use connections to communicate
with each other by sending packets subclassed from
\textsf{eq::net::NodePacket}.

Nodes, like connections, are either in the closed, listening or
connected state. A listening node uses one or more listening connections
in conjunction with a connection set to wait for data or connection
requests. For each connection request, a connection and node are created
to represent this node on the local machine. This node is in the
connected state, and its connection is added to the connection set of
the listening node.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netNode.pdf}
  {\caption{\small\label{fNetNode}Communication between two nodes}}
\end{wrapfigure}
\fig{fNetNode} shows the communication between two nodes. The nodes in
the listening state also have a listening connection, which is not shown
in the Figure for complexity reasons.

When the remote node sends a packet, the listening node receives the
packet and dispatches it from the receiver thread using the method
\textsf{dispatchCommand}. The default implementation knows how to
dispatch packets of type node, session or object. Applications can
define custom data types for packets, and then have to extend
\textsf{dispatchCommand} to handle these custom data types.

Node packets are directly dispatched using
\textsf{Base::dispatchCommand}. For session and object packets, the
appropriate session is found and \textsf{Session::dispatchCommand} will
dispatch the packet.

If \textsf{dispatchCommand} returns false, the packet will be
redispatched at a later time. This is used if an object has not been
mapped locally, and therefore the packet could not be dispatched.

If an application wants to extend communication on the node level, it
can either define its own datatype for packets, or to define custom node
packets. Packets with a custom datatype are handled by overriding
\textsf{Node::dispatchCommand}. Custom node packets can use any command
greater than \textsf{CMD\_NODE\_CUSTOM} as the packet command. By
registering a \textsf{eq::net::CommandFunc} for these commands, the
\textsf{eq::net::Base} dispatch and invoke mechanism can be used
automatically.

\subsubsection{Session}

The \textsf{eq::net::Session} gathers provides higher-level
functionality to a set of nodes. The master instance of a session is
registered to a node, which makes this node the session server and
assigns a node-unique identifier to the session. All other nodes map
their instance of the session to their local listening node using the
session's identifier. One given node can participate in any number of
sessions, both as master and slave. This allows the flexible usage of
sessions to group different aspects of an application.

A session provides the facilitate to allocate session-unique
identifiers. These identifiers can be used by the application to
uniquely address data within the cluster. They are also used by the
session to address distributed objects, which are described in the
following section.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netSession.pdf}
  {\caption{\small\label{fNetSession}Communication between two Sessions}}
\end{wrapfigure}
Sessions instances can communicate to each other directly using an
\textsf{eq::net::Ses\-sion\-Packet}, which is derived from the
\textsf{NodePacket} and extends it by the session identifier. The
session ID is used by the node to dispatch the packet to the session.

The session, similar to the node, uses the methods
\textsf{dispatchCommand} and \textsf{invokeCommand} to dispatch and
execute packets using either the \textsf{eq::net::Base} class for
session packets or to dispatch and invoke object packets on the
appropriate \textsf{eq::net::Object}. Application developers can use
this mechanism by creating custom session packets with a command greater
than \textsf{CMD\_SESSION\_CUSTOM} and registering a command function
for these commands.

\subsubsection{Object}

Distributed objects provide powerful, object-oriented data
distribution for C++ objects. Their functionality and an example use
case for parallel rendering has been described in \sref{sNetObject}.

The master instance of an \textsf{eq::net::Object} is registered with a
session, which assigns a session-unique identifier to the object. Using
this identifier, other nodes can map their slave instance of the object
to their local proxy of the session.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
  {\caption{\small\label{fNetObject}Communication between two objects}}
\end{wrapfigure}
Objects can send command packets to each other by subclassing
\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
\textsf{SessionPacket}. The object packet extends the session packet by
the object identifier and an object instance identifier.

Multiple instances of the same object (identifier) can be mapped on the
same node and session simultaneously. Each object instance has a
node-unique instance identifier. When the instance ID is set to
\textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
object instance registered to the the session with the correct object
ID. Multiple object instance per node are used when different threads on
the node need to process a different version of the same object.

Distributed objects provide data synchronization functionality, as
described in \sref{sNetObject}. Additionally, applications can send
command packets by creating custom object packets with a command greater
than \textsf{CMD\_OBJECT\_CUSTOM}.

Distributed object communication relies on the lower node and session
communication layers, as shown in \fig{fNetObject}.

\subsubsection{Barrier}

The \textsf{eq::net::Barrier} provides a networked barrier primitive. It
is an \textsf{eq::net::Object} used by Equalizer for software swap
barrier support, but it can be used as a generic barrier in application
code.

The barrier uses both the data distribution for synchronizing its data,
as well as custom command packets to implement the barrier logics. A
barrier is a versioned object. Each version can have a different height,
and enter request are automatically grouped together by version.
